Adaptive Waveform Inversion for Transmitted Wave Data

### 2024-11-11 repro/papers/awi/paper.tex
final, goes to arxiv

The arguments presented here for AWI are refinements of those given by
\cite{HuangSymes:Geo17} for Matched Source Waveform Inversion (MSWI),
a similar modification of FWI
\cite[]{HuangSymes2015SEG,HuangSymes:Geo17}. which were in turn based
on ideas presented in Song's thesis \cite[]{Song:94c}. Since it
involves no additional effort, I state results for MSWI in parallel
with those for AWI.

@@

  \begin{itemize}
  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
    L^2(\bR)$;
  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
    \lambda_*\le 1$ depending on
    $M, X_{sr}$ so that
    \begin{itemize}
    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$.
    \end{itemize}
  \end{itemize}
  Choose $r > 0$ and assume that
  \begin{equation}
    \label{eqn:lamsig}
    \sigma = r \lambda.
  \end{equation}.
  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
  \rightarrow \bR^+$ with $C_* \le
  W \le C^*$ so that for
  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{itemize}
  \item[2. ] for $0 <  \lambda \le 1$, 
    \begin{equation}
      \label{eqn:mswitt}
      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} W[\kappa^*,\kappa]^{\frac{1}{2}}(\bx_s,\bx_r)|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
      \lambda^{\frac{1}{2}};
    \end{equation}
  \item[3. ]for sufficiently small $\lambda$, 
    \begin{equation}
      \label{eqn:awitt}
      \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
        |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
    \end{equation}
  \end{itemize}
\end{theorem}
\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
2b show that for this to be the case, the regularization parameter $\sigma$
must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
would clearly suffice, however the argument lis slightly cleaner if
$\sigma$ is simply held proportional to $\lambda$ as $\lambda
\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.

@@

This paper has explained the behaviour of AWI (and MSWI) applied to
acoustic transmission data. For data exhibiting a single smooth
wavefront, these formulatios produce objective function that converge
to versions of travel time mean-square error, in the limit of hgh
data frequency. For

The conclusions of this study depend on a number of limitations on the
acoustic model of wave propagation and the formulation of the related
inverse problem. The principal constraints are:
\begin{itemize}
\item energy sources are modeled as isotropic point radiators, with
  known time dependence vanishing outside a finite time window;
\item mechanical parameter fields (coefficients in the acoustic wave
  equation) are smooth, that is, infinitely differentiable as
  functions of position, and the material density is known and
  position-independent: only the bulk modulus (or wave velocity) is
  to be determined;
\item waves propagate throughout Euclidean space-time - that is, no
  physical boundaries are modeled;
\item the data of the inverse problem are the restrictions of the
  pressure field to a finite number of points in space (receiver
  positions), for a finite number of source positions and all time;
\item local energy decays exponentially - that is, the pressure and
  all of its derivatives tend to zero at an exponential rate as $t
  \rightarrow \infty$,
  uniformly in any bounded spatial domain.
\item only one ray of geometric
  acoustics connects each source-receiver location pair (single arrivals)
\end{itemize}

Some of these constraints
could certainly be relaxed. For instance, exponential decay of local
energy implies that the finite time duration of actual measured data
can be treated as a perturbation of the infinite time data setting
posed here. The conclusions of this paper likely hold in some form
even for non-smooth sets of models, provided that the data for such models is
dominated by ballistic transmitted waves. The exception is the single-arrival hypothesis: it is
essential, as without it, the link between the travel time and
extended waveform inversion is lost.

This disconnect is a fearure of the adaptive filter construction, not of a particular objective function
(MSWI, AWI,...) formulated in terms of it. The introduction of the
adaptive filter $u$ in the filtered data prediction $F[\kappa]*u$ is equivalent
to permitting the source wavelet to depend on the receiver location,
which in the basic physical model of point source acoustics it does
not. Therefore this construction {\em extends} basic physical modeling
of acoustic wave transmission. \cite{HuangSymes:SEG15} call this is the {\em source-receiver
  extension}.  This ``crosstalk'' ambiguity analyzed in the previous
section afflicts all approaches based on the source-receiver
extension.

A number of authors have presented numerical illustrations of the
divergence between source-receiver extension methods and traveltime
tomography, and some have proposed remedies. \cite{Symes:94c}
illustrated the crosstalk phenomenon for a version of MSWI. Crosswell
source-receiver geometry often generates multiple arrivals,
particularly near low-velocity sedimentary layers that are natural
targets of seismic exploration and act as wave guides.
\cite{Plessix:00} successfully applied a version of MSWI to crosswell
waveform inversion by removing traces with close source and receiver
depths, thus leaving only data that tended to be dominated by single
arrivals. \cite{HuangSymes:Geo17} and \cite{SymesChenMinkoff:24}
illustrate the failure of the version of MSWI reported here to avoid
cycle-skipping in inversion of strongly refracting models, using
frequency-domain implementation of the basic definition
\ref{eqn:mswidefmain} and time-domain implementation of the reduced
penalty function \ref{eqn:mswipenred}, respectively.
\cite{Yongetal:GJI23} have shown that precisely the same issue arises
in AWI, and have proposed a remedy by localizing the definition of the
objective to short time intervals, within which multiple wavefront
arrrivals are unlikely to appear. \cite{Symes:23} constructs an
extension based on surface sources, as opposed to the point source -
point receiver extension which is the basis of MSWI and AWI. This
surface source extension separates events by infinitesimal slope,
information not available to the source-receiver extension.

Many open questions remain about MSWI, AWI, and other extension-based
approaches to transmitted wave data inversion. In particular, 
the analysis presented here has nothing to say about AWI applied to
reflected wave data. \cite{Warner:16} and others have
suggested that AWI is effective in that context. If so, the underlying
mechanism remains to be explained. The application of extended model
inversion to of elastodynamic transmission data remains to be investigated.

>>>>

The arguments presented here for AWI are refinements of those given by
\cite{HuangSymes:Geo17} for Matched Source Waveform Inversion (MSWI),
a similar modification of FWI
\cite[]{HuangSymes2015SEG,HuangSymes:Geo17}. which were in turn based
on ideas presented in Song's thesis \cite[]{Song:94}. Since it
involves no additional effort, I state results for MSWI in parallel
with those for AWI.

@@

  \begin{itemize}
  \item[1. ] for $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$, and
    $\sigma, \lambda >0$, $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
    L^2(\bR)$;
  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
    \lambda_*\le 1$ depending on
    $M, X_{sr}$ so that
    \begin{itemize}
    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$.
    \end{itemize}
  \end{itemize}
  Choose $r > 0$ and assume that
  \begin{equation}
    \label{eqn:lamsig}
    \sigma = r \lambda.
  \end{equation}.
  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
  \rightarrow \bR^+$ with $C_* \le
  W \le C^*$ so that for
  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{itemize}
  \item[2. ] for $0 <  \lambda \le 1$, 
    \begin{equation}
      \label{eqn:mswitt}
      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} W[\kappa^*,\kappa]^{\frac{1}{2}}(\bx_s,\bx_r)|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
      \lambda^{\frac{1}{2}};
    \end{equation}
  \item[3. ]for sufficiently small $\lambda$, 
    \begin{equation}
      \label{eqn:awitt}
      \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
        |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
    \end{equation}
  \end{itemize}
\end{theorem}
\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
2b show that for this to be the case, the regularization parameter $\sigma$
must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
would clearly suffice, however the argument lis slightly cleaner if
$\sigma$ is simply held proportional to $\lambda$ as $\lambda
\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.

@@

This paper has explained the behaviour of AWI (and MSWI) applied to
acoustic transmission through smoothly varying material models. For
isotropic point source/receiver data exhibiting a single smooth
wavefront, these formulations produce objective functions that converge
to versions of travel time mean-square error, in the limit of hgh
data frequency.

%The conclusions of this study depend on a number of limitations on the
%acoustic model of wave propagation and the formulation of the related
%inverse problem. The principal constraints are:
%\begin{itemize}
%\item energy sources are modeled as isotropic point radiators, with
%  known time dependence vanishing outside a finite time window;
%\item mechanical parameter fields (coefficients in the acoustic wave
%  equation) are smooth, that is, infinitely differentiable as
%  functions of position, and the material density is known and
%  position-independent: only the bulk modulus (or wave velocity) is
%  to be determined;
%\item waves propagate throughout Euclidean space-time - that is, no
%  physical boundaries are modeled;
%\item the data of the inverse problem are the restrictions of the
%  pressure field to a finite number of points in space (receiver
%  positions), for a finite number of source positions and all time;
%\item local energy decays exponentially - that is, the pressure and
%  all of its derivatives tend to zero at an exponential rate as $t
%  \rightarrow \infty$,
%  uniformly in any bounded spatial domain.
%\item only one ray of geometric
%  acoustics connects each source-receiver location pair (single arrivals)
%\end{itemize}

Some of the assumptions leading to this conclusion 
could certainly be relaxed. For instance, neither absence of boundary effects
nor exponential decay of local energy are essential. Instead
the finite time duration of actual measured data, together with more
elaborate ray geometry assumptions, can replace the infinite time data setting
posed here,. The conclusions of this paper likely hold in some form
even for non-smooth material parameter variation, provided that the data for such models is
dominated by ballistic transmitted waves. Similar conclusions should
hold for multipole point sources and
receivers, modeling non-isotropic experimental devices that are small on the
wavelength scale.

The exception is the single-arrival hypothesis: it is
essential, as without it, the link between the travel time and
extended waveform inversion is lost.
%This disconnect is a fearure of the adaptive filter construction, not of a particular objective function
%(MSWI, AWI,...) formulated in terms of it. The introduction of the
%adaptive filter $u$ in the filtered data prediction $F[\kappa]*u$ is equivalent
%to permitting the source wavelet to depend on the receiver location,
%which in the basic physical model of point source acoustics it does
%not. Therefore this construction {\em extends} basic physical modeling
%of acoustic wave transmission. \cite{HuangSymes:SEG15} call this is the {\em source-receiver
%  extension}.  This ``crosstalk'' ambiguity analyzed in the previous
%section afflicts all approaches based on the source-receiver
%extension.
A number of authors have illustrated this disconnect via numerical
example. The first such examples for a version of MSWI appeared in
\cite{Symes:94c}. Crosswell seismic surveys often generate multiple
arrivals near low-velocity sedimentary layers that are natural targets
of seismic exploration and act as wave guides. \cite{Plessix:00}
successfully applied a version of MSWI to crosswell waveform inversion
by removing traces with close source and receiver depths, thus leaving
only data that tended to be dominated by single arrivals.
\cite{HuangSymes:Geo17} and \cite{SymesChenMinkoff:24} illustrate the
failure of the version of MSWI reported here to avoid cycle-skipping
in inversion of strongly refracting models, using frequency-domain
implementation of the basic definition \ref{eqn:mswidefmain} and
time-domain implementation of the reduced penalty function
\ref{eqn:mswipenred}, respectively. \cite{Yongetal:GJI23} show that
precisely the same shortcoming afflicts AWI.

Overcoming this limitation requires substantial change in the
extension of basic acoustic modeling underlying AWI. For example,
\cite{Yongetal:GJI23} have proposed localizing the definition of the
AWI objective to short time intervals, within which multiple wavefront
arrrivals are unlikely to appear, and illustrated the effectiveness of
this approach by numerical example. \cite{Symes:23} constructs an
extension based on surface sources, as opposed to the trace-by-trace
extension which is the basis of MSWI and AWI. This surface source
extension separates events by infinitesimal slope, information not
available in the trace-by-trace construction..

The analysis presented here has nothing to say about AWI applied to
reflected wave data. \cite{Warner:16} and others have
suggested that AWI is effective in that context. If so, the underlying
mechanism remains to be explained.

Linear elasticity is a much better model for mechanical waves
propagating in some human tissue (notably bone) and the solid earth
than is acoustodynamics. Least-squares FWI of elastodynamic
transmission data also tends to stagnate due to cycle-skipping, and it
is tempting to think that extension approaches such as AWI should
ameliorate that obstacle. It will be necessary extract wave speed
informaiton in the presence of potential interference of cross-talk
between wave modes, similar to cross-talk between multiple arrivals in
the acoustic case.

Finally, virtually all materials supporting wave propagation also
attenuate energy through various physical mechanisms, which act per
cycle and therefore suppress high frequency components. Analysis such
as that presented here, relying on the infinite-frequency limit
(geometric optics), is {\em a priori} impossible for such attenuative
wave physics without fundamental
reformulation..
